---
title: "Project 1"
author: "Jordan Klein"
date: "2025-10-17"
output: html_document
---

---------------------------------
## YOUTUBE Link
https://www.youtube.com/watch?v=xUyYTIhuPhA
---------------------------------
## INIT
Clear the environment.
Import libraries and data.
Set up labelled and ordered factors.
---------------------------------
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Fresh start
rm(list = ls())

#Libraries
library(tidyverse)
library(naniar)
library(class)
library(caret)
library(e1071)
library(patchwork)
library(ggthemes)
library(gt)
library(scales)
library(rlang)
library(themis)
library(ggridges)
library(smotefamily)

#Load Data
data <- read.csv(file.choose())

#Set Factors
data$Attrition <- as.factor(data$Attrition)
data$BusinessTravel <- factor(
  data$BusinessTravel,
  levels = c("Non-Travel", "Travel_Rarely", "Travel_Frequently"),
  ordered = TRUE
)
data$Department <- as.factor(data$Department)
data$Education <- factor(
  data$Education,
  labels = c("Below College", "College", "Bachelor", "Master", "Doctor"),
  ordered = TRUE
)
data$EducationField <- as.factor(data$EducationField)
data$Gender <- as.factor(data$Gender)
data$JobInvolvement <- factor(
  data$JobInvolvement,
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
data$JobLevel <- factor(
  data$JobLevel,
  labels = c("Entry", "Junior", "Mid", "Senior", "Executive"),
  ordered = TRUE
)
data$JobRole <- as.factor(data$JobRole)
data$MaritalStatus <- as.factor(data$MaritalStatus)
data$Over18 <- as.factor(data$Over18)
data$OverTime <- as.factor(data$OverTime)
data$PerformanceRating <- factor(
  data$PerformanceRating,
  levels = c(1, 2, 3, 4),
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
data$RelationshipSatisfaction <- factor(
  data$RelationshipSatisfaction,
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
data$StockOptionLevel <- factor(
  data$StockOptionLevel,
  labels = c("None", "Low", "Medium", "High"),
  ordered = TRUE
)
data$WorkLifeBalance <- factor(
  data$StockOptionLevel,
  labels = c("Bad", "Good", "Better", "Best"),
  ordered = TRUE
)

# Color Scheme: Solarized

base0 = "#839496"
base1 = "#93a1a1"
base2 = "#eee8d5"
base3 = "#fdf6e3"
yellow = "#b58900"
orange = "#cb4b16"
red = "#dc322f"
magenta = "#d33682"
violet = "#6c71c4"
blue = "#268bd2"
cyan = "#2aa198"
green = "#859900"

# Define Column Groupints

# `Know Your Employee`
KYE <- c(
  "Age",
  "Department",
  "Education",
  "EducationField",
  "Gender",
  "MaritalStatus",
  "NumCompaniesWorked"
)

QualityOfLife <- c(
  "BusinessTravel",
  "DistanceFromHome",
  "EnvironmentSatisfaction",
  "JobInvolvement",
  "JobSatisfaction",
  "OverTime",
  "RelationshipSatisfaction",
  "WorkLifeBalance"
)

Reward <- c(
  "HourlyRate",
  "MonthlyIncome",
  "MonthlyRate",
  "StockOptionLevel",
  "PercentSalaryHike"
)

CareerProgression <- c(
  "JobRole",
  "JobLevel",
  "TotalWorkingYears",
  "TrainingTimesLastYear",
  "YearsAtCompany",
  "YearsInCurrentRole",
  "YearsSinceLastPromotion",
  "YearsWithCurrManager"
)

data %>% 
  select(JobLevel, StockOptionLevel, Attrition) %>% 
  filter(
    JobLevel == "Entry",
    StockOptionLevel == "High"
  ) 
```

---------------------------------
## MODEL HELPER FUNCTIONS

You must run this code block before evaluating the models on test data.

The helper functions are categorized by their role in the model development \
pipeline:
  - Preprocessing
  - Inference
  - Evaluation

---------------------------------
```{r}
# ---------------------------------
# PREPROCESSING
# Contains function for scaling, splitting, and slicing data for modeling.
# ---------------------------------

train_test_split <- function( df, train_size = 0.7) {
  
  # Randomly shuffles and splits a data frame into training and testing sets.
  #
  # Args:
  #   - df: A data frame to be split.
  #   - train_size: A numeric value (between 0 and 1) indicating the proportion
  #       of data to include in the training set. Default is 0.7.
  #
  # Returns:
  #   splits: list containing two data frames:
  #     - train: The training subset of the data.
  #     - test:  The testing subset of the data.
  #
  
  # --- Randomly select train indices ---
  n <- nrow(df)
  train_indices <- sample(seq_len(n), size = floor(train_size * n))
  
  # --- Segment data using the indices ---
  train <- df[train_indices, , drop = FALSE]
  test  <- df[-train_indices, , drop = FALSE]
  
  return(list(train = train, test = test))
}

slice_train_test <- function(splits, features, label) {
  
  # Splits the training and testing data into feature and label sets.
  #
  # Args:
  #   - splits: A list containing 'train' and 'test' data frames.
  #   - features: A character vector of column names to be used as features.
  #   - label: A string specifying the name of the label (target) column.
  #
  # Returns:
  #   A list containing:
  #     - train.X: Training features (data frame)
  #     - train.Y: Training labels (vector)
  #     - test.X: Testing features (data frame)
  #     - test.Y: Testing labels (vector)
  
  train <- splits$train
  test  <- splits$test
  
  train.X <- train[ , features, drop = FALSE]
  train.Y <- as.factor(train[[label]])
  test.X  <- test[ , features, drop = FALSE]
  test.Y <- as.factor(test[[label]])
  
  return(list(train = train, test = test, train.X = train.X, train.Y = train.Y, test.X = test.X, test.Y = test.Y))
}

apply_scaling <- function(splits) {
  # Applies heuristic scaling to feature sets, fitting only on training data.
  #
  # Args:
  #   - splits: A list containing 'train.X' and 'test.X' data frames.
  #
  # Returns:
  #   The same 'splits' list with scaled 'train.X' and 'test.X'.
  
  # --- Helper to normalize numeric columns (0–1) ---
  normalize <- function(x, ref_x) {
    rng <- range(ref_x, na.rm = TRUE)
    if (diff(rng) == 0) return(rep(0, length(x)))
    (x - rng[1]) / diff(rng)
  }
  
  train.X <- splits$train.X
  test.X  <- splits$test.X
  
  # --- Identify column types based on training data ---
  num_cols <- names(train.X)[sapply(train.X, is.numeric)]
  ord_cols <- names(train.X)[sapply(train.X, is.ordered)]
  fac_cols <- names(train.X)[sapply(train.X, function(x) is.factor(x) && !is.ordered(x))]
  
  train_scaled <- train.X
  test_scaled  <- test.X
  
  # --- Scale numeric features ---
  if (length(num_cols) > 0) {
    train_scaled[num_cols] <- lapply(train.X[num_cols], normalize, ref_x = train.X[num_cols])
    test_scaled[num_cols]  <- Map(normalize, test.X[num_cols], train.X[num_cols])
  }
  
  # --- Scale ordered factors (convert to numeric 0–1 ranks) ---
  if (length(ord_cols) > 0) {
    to_scaled_ord <- function(x, ref_x) {
      as.numeric(x) / max(as.numeric(ref_x), na.rm = TRUE)
    }
    train_scaled[ord_cols] <- Map(to_scaled_ord, train.X[ord_cols], train.X[ord_cols])
    test_scaled[ord_cols]  <- Map(to_scaled_ord, test.X[ord_cols], train.X[ord_cols])
  }
  
  # --- One-hot encode unordered factors using training levels ---
  if (length(fac_cols) > 0) {
    dummies <- caret::dummyVars(~ ., data = train.X[fac_cols])
    train_fac <- as.data.frame(predict(dummies, newdata = train.X[fac_cols]))
    test_fac  <- as.data.frame(predict(dummies, newdata = test.X[fac_cols]))
    
    train_scaled <- bind_cols(select(train_scaled, -all_of(fac_cols)), train_fac)
    test_scaled  <- bind_cols(select(test_scaled, -all_of(fac_cols)), test_fac)
  }
  
  splits$train.X <- train_scaled
  splits$test.X  <- test_scaled
  
  return(splits)
}

apply_smote <- function(splits, perc_over = 200, perc_under = 200, k = 5, label = "Attrition") {
  # Applies SMOTE balancing to the training data in 'splits'.
  #
  # Args:
  #   - splits: A list with elements train.X and train.Y.
  #   - perc_over: Percentage of over-sampling for minority classes.
  #   - perc_under: Percentage of under-sampling for majority classes (not used here).
  #   - k: Number of nearest neighbors for SMOTE.
  #   - label: Name of the target column.
  #
  # Returns:
  #   Updated 'splits' list with balanced train.X and train.Y.
  
  # Combine features and label into one data frame for SMOTE
  train_data <- data.frame(splits$train.X, target = splits$train.Y)
  
  # Count samples in the smallest class
  class_counts <- table(train_data$target)
  min_class_size <- min(class_counts)
  
  # Ensure K < min_class_size
  k_safe <- min(k, min_class_size - 1)
  if (k_safe < 1) k_safe <- 1
  
  # Apply SMOTE
  smote_result <- smotefamily::SMOTE(X = train_data[, -ncol(train_data)],
                                     target = train_data$target,
                                     K = k_safe,
                                     dup_size = perc_over / 100)
  
  balanced <- smote_result$data
  # The last column is the target variable created by SMOTE
  balanced[[ncol(balanced)]] <- as.factor(balanced[[ncol(balanced)]])
  
  # Split back into X and Y
  splits$train.X <- balanced[, -ncol(balanced), drop = FALSE]
  splits$train.Y <- balanced[[ncol(balanced)]]
  
  return(splits)
}

# ---------------------------------
# FEATURE ENGINEERING
# Add features to improve model performance.
# ---------------------------------

add_joblevel_group <- function(splits) {
  
  # Adds a JobLevelGroup column to both train and test sets.

  
  for (name in c("train", "test")) {
    splits[[name]] <- splits[[name]] %>%
      mutate(
        JobLevelGroup = case_when(
          JobLevel %in% c("Entry", "Junior") ~ "Low",
          JobLevel %in% c("Mid", "Senior") ~ "Middle",
          TRUE ~ "High"
        ),
        JobLevelGroup = factor(JobLevelGroup, levels = c("Low", "Middle", "High"))
      )
  }
  
  return(splits)
}

add_department_group <- function(splits) {
  
  # Adds a DepartmentGroup column to both train and test sets.
  
  for (name in c("train", "test")) {
    splits[[name]] <- splits[[name]] %>%
      mutate(
        DepartmentGroup = case_when(
          Department == "Sales" ~ "Sales",
          TRUE ~ "NotSales"
        ),
        DepartmentGroup = as.factor(DepartmentGroup)
      )
  }
  
  return (splits)
}

add_income_residuals <- function(splits) {
  
  # Calculates residuals, and adds a categorical label: Below Expected, Around Expected, Above Expected
  
  x <- "YearsAtCompany"
  y <- "MonthlyIncome"
  groups <- c("JobLevel", "Department")
  
  train <- splits$train
  test <- splits$test
  
  train <- train %>% mutate(IncomeResidual = NA_real_)
  test  <- test  %>% mutate(IncomeResidual = NA_real_)
  
  group_combos <- train %>% distinct(across(all_of(groups)))
  
  for (i in seq_len(nrow(group_combos))) {
    grp_vals <- group_combos[i, ]
    
    # Select rows belonging to this group
    train_grp_idx <- which(rowSums(sapply(groups, function(col) train[[col]] == grp_vals[[col]])) == length(groups))
    train_grp <- train[train_grp_idx, ]
    
    if (nrow(train_grp) < 2) next
    
    # Skip if YearsAtCompany is constant or all NA
    if (length(unique(train_grp[[x]][!is.na(train_grp[[x]])])) < 2) next
    
    model <- lm(reformulate(x, y), data = train_grp)
    
    # Residuals for train
    train$IncomeResidual[train_grp_idx] <- train_grp[[y]] - predict(model, newdata = train_grp)
    
    # Residuals for test (only for groups seen in train)
    test_grp_idx <- which(rowSums(sapply(groups, function(col) test[[col]] == grp_vals[[col]])) == length(groups))
    if (length(test_grp_idx) == 0) next
    
    preds <- predict(model, newdata = test[test_grp_idx, ])
    test$IncomeResidual[test_grp_idx] <- test[test_grp_idx, y] - preds
  }
  
  # Optionally fill remaining NAs with mean residual (centered at 0)
  train$IncomeResidual[is.na(train$IncomeResidual)] <- 0
  test$IncomeResidual[is.na(test$IncomeResidual)] <- 0
  
  splits$train <- train
  splits$test <- test
  return(splits)
}

add_education_group <-function(splits) {
  
  # Adds a EducationGroup column to both train and test sets.
  
  for (name in c("train", "test")) {
    splits[[name]] <- splits[[name]] %>%
      mutate(
        EducationGroup = case_when(
          JobLevel %in% c("Below College", "College") ~ "Low",
          JobLevel %in% c("Bachelor") ~ "Middle",
          TRUE ~ "High"
        ),
        JobLevelGroup = factor(JobLevelGroup, levels = c("Low", "Middle", "High"))
      )
  }
  
  return(splits)
}

   
engineer_features <- function(splits, choices = c()) {
  
  # Ties together all feature engineering options
  
  
  if ("JobLevelGroup" %in% choices) {
    splits <- add_joblevel_group(splits)
  }
  
  if ("DepartmentGroup" %in% choices) {
    splits <- add_department_group(splits)
  }
  
  if ("IncomeResidual" %in% choices) {
    splits <- add_income_residuals(splits)
  }
  
  if ("EducationGroup" %in% choices) {
    splits <- add_education_group(splits)
  }
  
  return (splits)
}



# ---------------------------------
# INFERENCE
# Given training data, has functions for making predictions with
# Naive Bayes and K-NN.
# ---------------------------------

get_knn_predictions <- function(splits, k=NULL, balance = FALSE) {
  
  # Generates predictions using the K-Nearest Neighbors (KNN) algorithm.
  #
  # Args:
  #   - splits: A list containing training and testing data components:
  #       - train.X — dataframe of training features
  #       - test.X  — dataframe of testing features
  #       - train.Y — vector of training labels
  #   - k: Integer specifying the number of nearest neighbors to use for prediction.
  #
  # Returns:
  #   A vector of predicted class labels for the test set.
    splits <- apply_scaling(splits)
  
  if (balance) {
    splits <- apply_smote(splits)
    }
  
  if(is.null(k)) {k <- floor(max(3, sqrt(nrow(splits$train))))}
  
  predictions <- class::knn(
    train = splits$train.X,
    test = splits$test.X,
    cl = splits$train.Y,
    k=k
    )
    return (predictions)
}

get_naivebayes_predictions <- function(splits) {
  
  # Generates predictions using Naive Bayes.
  #
  # Args:
  #   - splits: A list containing training and testing data components:
  #       - train.X — dataframe of training features
  #       - test.X  — dataframe of testing features
  #       - train.Y — vector of training labels
  #   - k: Integer specifying the number of nearest neighbors to use for prediction.
  #
  # Returns:
  #   A vector of predicted class labels for the test set.
  
  model <- e1071::naiveBayes(
    x = splits$train.X, 
    y = splits$train.Y
    )
  
  predictions <- predict(
    object = model, 
    newdata = splits$test.X
    )
  
  return(predictions)
}

run_model <- function(splits, model_class, k = NULL, balance = FALSE) {
  
  if (model_class == "knn") {
    predictions <- get_knn_predictions(splits, k=k, balance = balance)
  } else if (model_class == "naivebayes") {
    predictions <- get_naivebayes_predictions(splits)
  }
  
  
  return(predictions)
}

get_predictions <- function(
  splits,
  model_class = "naivebayes",
  k = NULL,
  group_col = NULL,
  label = "Attrition",
  balance = FALSE
) {
  # Runs inference either on the entire dataset or grouped by a factor column.
  #
  # Args:
  #   - splits: List containing train/test data and features/labels.
  #   - model_class: "knn" or "naivebayes".
  #   - k: Number of neighbors (for KNN only).
  #   - group_col: Optional string; name of column to group by.
  #
  # Returns:
  #   - A vector of predictions in the same order as the original test set.
  
  # --- CASE 1: No grouping, evaluate normally ---
  if (is.null(group_col)) {
    preds <- run_model(splits, model_class=model_class, balance = balance)
    return(preds)
  }
  
  # --- CASE 2: Grouped inference ---
  train <- splits$train
  test <- splits$test
  
  # Store predictions in same order as test
  preds <- factor(rep(NA, nrow(test)), levels = levels(splits$test.Y))
  
  # Iterate over each group present in test data
  for (grp in unique(test[[group_col]])) {
    train_grp <- train[train[[group_col]] == grp, , drop = FALSE]
    test_grp <- test[test[[group_col]] == grp, , drop = FALSE]
    
    if (nrow(train_grp) == 0 || nrow(test_grp) == 0) {
      next  # skip groups missing from either set
    }
    
    # Slice features and labels
    sub_splits <- slice_train_test(
      splits = list(train = train_grp, test = test_grp),
      features = colnames(splits$train.X),
      label = label
    )
    
    # Run inference for this group
    grp_preds <- run_model(sub_splits, model_class=model_class, balance=balance)
    
    # Place predictions back into correct rows
    preds[rownames(test) %in% rownames(test_grp)] <- grp_preds
  }
  
  return(preds)
}



# ---------------------------------
# Evaluation
# 
# Functions for projecting the cost of attrition and intervention, plotting
# and rendering evaluation summaries, and apploying the procedure over a series
# of random splits.
# ---------------------------------

get_eval_df <- function(test, predictions, attrition_multiplier = 12, intervention_multiplier = 3)   {
  
  # Prepares an evaluation DataFrame to assess the cost impact of attrition 
  # and interventions, and to evaluate specificity across departments and job levels.
  #
  # Args:
  #  - test: DataFrame containing the test data.
  #  - predictions: A vector of model predictions.
  #  - attrition_multiplier: Multiplier to compute attrition cost, defaults to 12 (Annual)
  #  - intervention_multiplier: Multiplier to computer intervention cost, defaults to 3 (Quarterly)
  #
  # Returns:
  #   eval: DataFrame containing evaluation records.
  

  eval <- test %>% 
    mutate(
      AttritionCost  = attrition_multiplier * MonthlyIncome,
      InterventionCost = intervention_multiplier * MonthlyIncome,
      Actual = Attrition,
      Predictions = predictions
    ) %>% 
    select(ID, Actual, Predictions, JobLevel, Department, AttritionCost, InterventionCost)
  
  return (eval)
}

summarise_cost_impact <- function(eval) {
  
  # Summarizes cost impact metrics from model evaluation results..
  #
  # Args:
  #   eval: DataFrame containing evaluation records, including cost data, 
  #         actual labels, and model predictions.
  #
  # Returns:
  #   DataFrame summarizing cost impact metrics, including:
  #     - Total Intervention Cost: The investment required to prevent attrition.
  #     - Cost of No Action: The estimated loss if no intervention is applied.
  #     - Unrealized Attrition Loss: The recoverable loss from correctly predicted attrition cases.
  #     - Realized Attrition Loss: The loss from employees who left but were not predicted to do so.
  #     - Potential Net Benefit: The net gain if all predicted employees are successfully retained.

  eval %>%
    mutate(
      # Define scenario flags
      TruePositive  = (Predictions == "Yes" & Actual == "Yes"),
      FalsePositive = (Predictions == "Yes" & Actual == "No"),
      FalseNegative = (Predictions == "No"  & Actual == "Yes")
    ) %>%
    summarise(
      TotalInterventionCost  = sum(InterventionCost[TruePositive | FalsePositive]),
      CostOfNoAction = sum(AttritionCost[TruePositive | FalseNegative]),
      UnrealisedAttritionLoss = sum(AttritionCost[TruePositive]),
      RealisedAttritionLoss   = sum(AttritionCost[FalseNegative])
    ) %>%
    mutate(
      PotentialNetBenefit = CostOfNoAction - RealisedAttritionLoss - TotalInterventionCost,
    )
    
}

summary_table_formatting <- function(gt_table) {
  
  # Global formatting instructions for gt tables.
  
  gt_table %>%
    # Bold column headers
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels(everything())
    ) %>%
    
    # Center and bold stubheads
    tab_style(
      style = cell_text(weight = "bold", align = "center"),
      locations = cells_stubhead()
    ) %>%
    
    # Add vertical dividing line between stub and data columns
    tab_style(
      style = cell_borders(sides = "right", color = base1, weight = px(1)),
      locations = cells_stub(rows = everything())
    ) %>%
    
    # Bold the final row
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_body(
        rows = nrow(gt_table$`_data`) 
      )
    ) %>% 
    
    # Global visual options
    tab_options(
      table.border.top.color = "white",
      table.border.bottom.color = "white",
      data_row.padding = px(6),
      table_body.hlines.color = base1,
      table.background.color = base3
    )
}

render_cost_impact_table <- function(results) {
  
  # Renders a presentation ready table from the results.
  
  metrics <- results$metrics
  impact <- results$impact
  
  # --- Assemble data frame ---
  metrics_df <- tibble::tibble(
    Metric = c(
      "Accuracy",
      "Sensitivity",
      "Specificity",
      "Balanced Acc",
      "F1",
      "Cost of No Action",
      "Cost of Intervention",
      "Realised Attrition",
      "Unrealised Attrition",
      "Potential Net Benefit"
    ),
    Value = c(
      percent(metrics$Accuracy, accuracy = 0.1),
      percent(metrics$Sensitivity, accuracy = 0.1),
      percent(metrics$Specificity, accuracy = 0.1),
      percent(metrics$`Balanced Accuracy`, accuracy = 0.1),
      number(metrics$F1, accuracy = 0.001),
      dollar(impact$CostOfNoAction, scale = 1, accuracy = 1),
      dollar(impact$TotalInterventionCost, scale = 1, accuracy = 1000),
      dollar(impact$RealisedAttritionLoss, accuracy = 1000),
      dollar(impact$UnrealisedAttritionLoss, accuracy = 1000),
      dollar(impact$PotentialNetBenefit, scale = 1, accuracy = 1000)
    )
  )
  
  # --- Build gt table ---
  metrics_table <- metrics_df %>%
    gt() %>%
    tab_header(
      title = md("**Model Metrics**")
    ) %>%
    cols_label(
      Metric = md("**Metric**"),
      Value = md("**Value**")
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels(everything())
    ) %>%
    tab_options(
      table.font.size = 14,
      table.width = pct(60),
      heading.title.font.size = 18,
      heading.align = "center"
    ) %>%  summary_table_formatting()
  
  return(metrics_table)
}

plot_confusion_matrix <- function(cm, relative = FALSE) {
  
  # Renders a presentation-ready confusion matrix.
  
  cm_df <- as.data.frame(cm)
  cm_df$Reference <- as.factor(cm_df$Reference)
  cm_df$Prediction <- as.factor(cm_df$Prediction)
  
  if (relative) {
    cm_df <- cm_df %>%
      group_by(Reference) %>%
      mutate(Freq = Freq / sum(Freq)) %>%
      ungroup()
  }
  
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) + 
    geom_tile(color = "white") + 
    geom_text(
      aes(
        label = if (relative) paste0(round(Freq * 100, 1), "%") else Freq
      ),
      size = 8, fontface = "bold"
    ) +
    scale_fill_gradient(low = base3, high = cyan) + 
    labs(
      title = if (relative) "CM of Attrition Predictons",
      x = "Actual", 
      y = "Predicted"
    ) +
    theme_solarized(base_size = 20) + 
    theme(legend.position = "none")
}

plot_sensitivity_matrix <- function(eval) {
  
  # Plots a heatmap that probes sensitivity across department and job level.
  
  p <- eval %>% 
    select(Actual, Predictions, JobLevel, Department) %>% 
    filter(Actual == "Yes") %>% 
    group_by(JobLevel, Department) %>% 
    summarise(Sensitivity = mean(Predictions == "Yes"), .groups = "drop") %>% 
    ggplot(aes(x = Department, y = JobLevel, fill = Sensitivity)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = base3, high = cyan) +
    geom_text(aes(label = percent(Sensitivity, accuracy = 1)), size = 8) +
    labs(
      title = "Sensitivity by Department and Job Level",
      x = "Department",
      y = "Job Level"
    ) +
    theme_solarized(base_size = 18) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  return (p)
}

check_for_missing_columns <- function(df, columns, split = "train") {

  # For sanity checking. Checks that all required columns are present in a dataset.

  
  # Identify which columns are missing from the dataframe
  missing <- setdiff(columns, colnames(df))
  
  # If any are missing, stop with an informative message
  if (length(missing) > 0) {
    stop(paste(
      "The following columns are missing from the", split, "dataset:",
      paste(missing, collapse = ", ")
    ))
  }

}

sanity_check <- function(
    splits, 
    features, 
    id = "ID", 
    label = "Attrition",
    group_col = NULL,
    value = c("MonthlyIncome", "Department", "JobLevel"),
    model_class = "naivebayes"
) {
  
  # Step 1 of the Model Eval Pipeline.
  # Checks for missing columns and ensures that the model class is valid.
  
    # Ensures that the model_class, features, label, and value columns are valid.
  
    # --- Validate model class ---
    valid_models <- c("knn", "naivebayes")
    if (!(tolower(model_class) %in% valid_models)) {
        warning(paste0(
            "Invalid model_class: '", model_class, 
            "'. Defaulting to 'naivebayes'. (Valid options: 'knn', 'naivebayes')"
        ))
        model_class <- "naivebayes"
    } else {
        model_class <- tolower(model_class)
    }
    
    # --- Check for missing columns in training data ---
    check_for_missing_columns(splits$train, c(features, id, label, group_col), split = "train")
    check_for_missing_columns(splits$test, c(features, label, group_col, value), split = "test")
}


model_eval <- function(
    data, 
    features,
    engineered_features = c(),
    id = "ID",
    label = "Attrition",
    model_class = "naivebayes",
    group_col = NULL,
    balance = FALSE,
    test = NULL,
    seed = 101, 
    train_size = 0.7,
    k = NULL,
    render_results = TRUE
) {
  
  # Main function that ties together preprocessing, prediction, and evaluation.
  #
  # Args:
  #   - data: The full dataset to sample from.
  #   - features: Character vector of feature column names.
  #   - id, label: Column names for ID and label.
  #   - model_class: Model type ("naivebayes", "knn", etc.).
  #   - test: If a test dataframe is input, will not generate splits, and will use test set instead
  #   - train_size: Proportion of data used for training each trial.
  #   - scale: Whether to apply scaling.
  #   - k: k value for k-NN. When NULL, defaults to sqrt(train size)
  #   - seed: Base random seed for reproducibility.
  #   - render_results: whether or not to produce plots
  #
  # Returns:
  #   A list containing:
  #     - eval: The eval dataset used to compute cost impact and sensitivity
  #.    - accuracy: The accuracy of model predictions
  #     - metrics: Model performance matrix
  #     - cm: A table of predictions vs actual labels
  #     - impact: Trial-level cost impacts
  #     - cm_plot: A rendered CM of the prediction results
  #     - impact_table: A table that summarizes metrics and  cost impact projections
  #     - sensitivity_matrix: A plot that probes the models sensitivity across subgroups
  
  set.seed(seed)
  
  
  # --- Split into train/test (if test not provided) ---
  if (is.null(test)) {
    splits <- train_test_split(data, train_size=train_size)
  } else {
    splits <- list(
      train = data,
      test = test
    )
  }
  splits <- engineer_features(splits, choices = engineered_features)
  sanity_check(splits, features, id=id, label=label)
  
  
  splits <- slice_train_test(splits, features, label)
  predictions <- get_predictions(
    splits, 
    model_class, 
    k=k,
    group_col = group_col,
    label = label,
    balance = balance
    )


  # --- Evaluate performance ---
  cm <- caret::confusionMatrix(predictions, splits$test.Y, positive = "Yes")
  metrics <- as.data.frame(t(cm$byClass))  # transpose so it becomes 1-row data frame
  metrics$Accuracy <- cm$overall["Accuracy"]
  eval <- get_eval_df(splits$test, predictions)

    # --- Output ---
  res <- list(
    eval = eval,
    confusion_matrix = cm$table,
    metrics = metrics,
    impact = summarise_cost_impact(eval)
  )

  if (render_results) {
    res$cm_plot <- plot_confusion_matrix(res$confusion_matrix)
    res$impact_table <- render_cost_impact_table(res)
    res$sensitivity_matrix <- plot_sensitivity_matrix(eval)
  }

  return(res)
}

post_process_impact <- function(all_impact, train_size, n_trials) {
  
  # Collapses cost_impact data, uses a weighted average to project company-wide 
  # cost impact
  
  # scaling factor: 1 / n_trials / test proportion
  scaling_factor <- 1 / n_trials / (1 - train_size)
  
  all_impact %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    mutate(across(where(is.numeric), ~ .x * scaling_factor))
}

render_cost_impact_table_circuit <- function(metrics_df, impact_df) {
  
  # Generates a presentation ready table from the combined cost impact data.
  
  tibble_df <- tibble::tibble(
    Metric = c(
      "Accuracy",
      "Sensitivity",
      "Specificity",
      "Balanced Acc",
      "F1",
      "Cost of No Action",
      "Cost of Intervention",
      "Realised Attrition",
      "Unrealised Attrition",
      "Potential Net Benefit"
    ),
    Value = c(
      percent(metrics_df$Accuracy, accuracy = 0.1),
      percent(metrics_df$Sensitivity, accuracy = 0.1),
      percent(metrics_df$Specificity, accuracy = 0.1),
      percent(metrics_df$`Balanced Accuracy`, accuracy = 0.1),
      number(metrics_df$F1, accuracy = 0.001),
      dollar(impact_df$CostOfNoAction, scale = 1, accuracy = 1000),
      dollar(impact_df$TotalInterventionCost, scale = 1, accuracy = 1000),
      dollar(impact_df$RealisedAttritionLoss, accuracy = 1000),
      dollar(impact_df$UnrealisedAttritionLoss, accuracy = 1000),
      dollar(impact_df$PotentialNetBenefit, scale = 1, accuracy = 1000)
    )
  )
  
  tibble_df %>% gt() %>% summary_table_formatting()
}

model_eval_circuit <- function(
    data,
    features,
    engineered_features = c(),
    n_trials = 100,
    id = "ID",
    label = "Attrition",
    model_class = "naivebayes",
    balance = FALSE,
    group_col = NULL,
    train_size = 0.7,
    k = NULL,
    seed = 101,
    render_results = TRUE,
    verbose = TRUE
) {
  
  # Repeated model evaluation over multiple random splits.
  #
  # Args:
  #   - data: The full dataset to sample from.
  #   - features: Character vector of feature column names.
  #   - n_trials: Number of random splits to evaluate.
  #   - id, label: Column names for ID and label.
  #   - model_class: Model type ("naivebayes", "knn", etc.).
  #   - train_size: Proportion of data used for training each trial.
  #   - scale: Whether to apply scaling.
  #   - k: k value for k-NN (if applicable).
  #   - seed: Base random seed for reproducibility.
  #   - verbose: Whether to print progress updates.
  #
  # Returns:
  #   A list containing:
  #     - eval: The eval dataset used to compute cost impact and sensitivity
  #     - metrics: Trial-level metrics (long-form)
  #     - cm: Combined confusion matrices (tidy form)
  #     - impact: Trial-level cost impacts (if available)
  #     - cm_plot: A rendered CM of the prediction results
  #     - impact_table: A table that summarizes metrics and  cost impact projections
  #     - sensitivity_matrix: A plot that probes the models sensitivity across subgroups

  
  set.seed(seed)
  
  all_eval <- tibble()
  all_metrics <- tibble()
  all_cm <- tibble()
  all_impact <- tibble()
  
  for (i in seq_len(n_trials)) {
    if (verbose && i %% 10 == 0) {
      message("Running trial ", i, "/", n_trials, "...")
    }
    
    res <- model_eval(
      data = data,
      features = features,
      engineered_features = engineered_features,
      id = id,
      label = label,
      balance = balance,
      group_col = group_col,
      model_class = model_class,
      train_size = train_size,
      k = k,
      render_results = TRUE,
      seed = seed + i
    )
    
    trial_eval <- res$eval
    trial_eval$trial <- 1
    all_eval <- bind_rows(all_eval, trial_eval)
    
    # --- Metrics ---
    trial_metrics <- as_tibble(res$metrics)
    trial_metrics$trial <- factor(i)
    all_metrics <- bind_rows(all_metrics, trial_metrics)
    
    # --- Confusion matrix ---
    trial_cm <- as.data.frame(res$confusion_matrix) %>%
      as_tibble() %>%
      mutate(trial = i)
    all_cm <- bind_rows(all_cm, trial_cm)
    
    # --- Impact summary ---
    trial_impact <- res$impact
    all_impact <- bind_rows(all_impact, trial_impact)
    
  }
  
  metrics_summary <- all_metrics %>% 
    summarise(across(where(is.numeric), mean, na.rm = TRUE))
  
  all_cm <- all_cm %>% 
    group_by(Prediction, Reference) %>% 
    summarize(Freq = sum(Freq), .groups = "drop"
              )
  
  all_impact <- post_process_impact(all_impact, train_size, n_trials)
  
  final_result <- list(
      eval = all_eval,
      metrics = all_metrics,
      cm = all_cm,
      impact = all_impact
    )
  
  if (render_results) {
    final_result$cm_plot <- plot_confusion_matrix(all_cm, relative = TRUE)
    final_result$impact_table <- render_cost_impact_table_circuit(metrics_summary, all_impact)
    final_result$sensitivity_matrix <- plot_sensitivity_matrix(all_eval)
  }
  
  return(final_result)
  
}

```
---------------------------------

## EDA FUNCTIONS
---------------------------------
```{r}

summarise_attrition <- function(
    df, 
    group_var, 
    arrange_col = AttritionRate, 
    descending = TRUE) {
  
  # Summarizes employee attrition by a specified grouping variable.
  #
  # Arguments:
  #   - df: A data frame containing employee data, including an 'Attrition' column.
  #   - group_var: The column to group by (e.g., Department, Team).
  #   - arrange_col: The column to sort the summary by (e.g., AttritionRate, N_Employees).
  #   - descending: Logical. If TRUE, sorts in descending order; otherwise ascending.
  #
  # Returns:
  #   - A summarized data frame with one row per group, including:
  #       * N_Employees: Number of employees in the group.
  #       * N_Attrited: Number of employees who have attrited ("Yes").
  #       * AttritionRate: Proportion of employees who have attrited.
  #     The summary also includes a final row labeled "TOTAL" with aggregated values across all groups.

  
  # Capture the grouping variable
  group_sym <- rlang::ensym(group_var)
  arrange_sym <- rlang::ensym(arrange_col)
  
  df_summary <- df %>%
    group_by(!!group_sym) %>% 
    summarize(
      N_Employees = n(),
      N_Attrited = sum(Attrition == "Yes", na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      AttritionRate = N_Attrited / N_Employees
    ) %>%
    ungroup()
  
  # Arrange by specified column
  if (descending) {
    df_summary <- df_summary %>% arrange(desc(!!arrange_sym))
  } else {
    df_summary <- df_summary %>% arrange(!!arrange_sym)
  }
  
  # Add total row
  df_summary %>%
    bind_rows(
      summarise(
        .,
        !!group_sym := "TOTAL",
        N_Employees = sum(N_Employees),
        N_Attrited = sum(N_Attrited),
        AttritionRate = sum(N_Attrited) / sum(N_Employees)
      )
    )
}

summarise_waste <- function(
    df, 
    group_var, 
    arrange_col = WasteRate, 
    descending = TRUE
    ) {
  
  # Summarizes payroll "waste" due to employee attrition by a specified grouping variable.
  #
  # Arguments:
  #   - df: A data frame containing employee data, including 'Attrition' and 'MonthlyIncome' columns.
  #   - group_var: The column to group by (e.g., Department, Team).
  #   - arrange_col: The column to sort the summary by (default is WasteRate).
  #   - descending: Logical. If TRUE, sorts in descending order; otherwise ascending.
  #
  # Returns:
  #   - A summarized data frame with one row per group, including:
  #       * Payroll: Total annual salary of all employees in the group.
  #       * Waste: Total annual salary lost due to employees who attrited ("Yes").
  #       * WasteRate: Proportion of payroll lost due to attrition (Waste / Payroll).
  #     The summary also includes a final row labeled "TOTAL" with aggregated values across all groups.
  
  # Capture the grouping variable
  group_sym <- rlang::ensym(group_var)
  arrange_sym <- rlang::ensym(arrange_col)
  
  df_summary <- df %>%
    # Compute annual salary
    mutate(AnnualSalary = 12 * MonthlyIncome) %>%
    
    # Keep only relevant columns
    select(!!group_sym, Attrition, AnnualSalary) %>%
    
    # Summarize by group and attrition
    group_by(!!group_sym, Attrition) %>%
    summarize(Total_Salary = sum(AnnualSalary), .groups = "drop") %>%
    
    # Pivot so Yes/No attrition becomes separate columns
    pivot_wider(
      names_from = Attrition,
      values_from = Total_Salary,
      values_fill = 0
    ) %>%
    
    # Compute waste metrics
    mutate(
      Waste = `Yes`,
      Payroll = `Yes` + `No`,
      WasteRate = Waste / Payroll
    ) %>%
    
    # Select final columns
    select(!!group_sym, Payroll, Waste, WasteRate) %>%
    ungroup()
  
  # Arrange by specified column
  if (descending) {
    df_summary <- df_summary %>% arrange(desc(!!arrange_sym))
  } else {
    df_summary <- df_summary %>% arrange(!!arrange_sym)
  }
  
  # Add total row
  df_summary %>%
    bind_rows(
      summarise(
        .,
        !!group_sym := "TOTAL",
        Payroll = sum(Payroll),
        Waste = sum(Waste),
        WasteRate = sum(Waste) / sum(Payroll)
      )
    )
}


render_attrition_table <- function(
    df, 
    group_var, 
    table_title = "Attrition Summary", 
    group_label = NULL
    ) {
  
  # Renders a presentation ready table that summarises attrion.
  
  group_col <- rlang::as_name(rlang::enquo(group_var))
  stub_label <- ifelse(is.null(group_label), group_col, group_label)
  
  df %>%
    gt(rowname_col = group_col) %>%
    tab_stubhead(label = stub_label) %>%
    cols_label(
      N_Employees = "Total Employees",
      N_Attrited = "Attrited Employees",
      AttritionRate = "Attrition Rate"
    ) %>%
    fmt_percent(columns = AttritionRate, decimals = 1) %>%
    fmt_number(columns = c(N_Employees, N_Attrited), decimals = 0) %>%
    tab_header(
      title = md(paste0("**", table_title, "**"))
    ) %>%
    summary_table_formatting()  
}

render_waste_table <- function(
    df, 
    group_var, 
    table_title = "Salary Waste Summary", 
    group_label = NULL) {
  
  # Renders a presentation ready table that summarises salary waste.

  group_col <- rlang::as_name(rlang::enquo(group_var))
  stub_label <- ifelse(is.null(group_label), group_col, group_label)
  
  df %>%
    ungroup() %>%
    gt(rowname_col = group_col) %>%
    tab_stubhead(label = stub_label) %>%
    cols_label(
      Payroll = "Total Payroll",
      Waste = "Attrition Waste",
      WasteRate = "Waste Rate"
    ) %>%
    fmt_currency(columns = c(Payroll, Waste), currency = "USD", decimals = 0) %>%
    fmt_percent(columns = WasteRate, decimals = 1) %>%
    tab_header(
      title = md(paste0("**", table_title, "**"))
    ) %>%
    summary_table_formatting()   
}


calc_residuals <- function(
    df, 
    x, 
    y, 
    groups = NULL, 
    threshold = 0.2) {
  
  # Calculates residuals from a simple linear model.
  # 
  # Fits a linear model using the given dependent (y) and independent (x) variables,
  # and appends the residuals as a new column in the dataset. If grouping variables
  # are provided, a separate model is fit within each group.
  #
  # Args:
  #   df: Data frame containing the variables.
  #   x: String. Name of the independent (predictor) variable.
  #   y: String. Name of the dependent (response) variable.
  #   groups: Optional character vector. Columns to group by for group-wise models.
  #   residuals_col: Optional string. Name of the output column for residuals. Defaults to "residual".
  #   threshold: For classifying residuals as Above or Below Expected
  #
  # Returns:
  #   The input data frame with an added column containing residuals.
  
  classify_residual <- function(resid, threshold) {
    case_when(
      resid > threshold ~ "Above Expected",
      resid < -threshold ~ "Below Expected",
      TRUE ~ "Around Expected"
    )
  }

  if (is.null(groups)) {
    model <- lm(reformulate(x, y), data = df)
    preds <- predict(model, df)
    df$residual <- (df[[y]] - preds) / df[[y]]
    df$residual_class <- classify_residual(df$residual, threshold)

  } else {
    df <- df %>%
      group_by(across(all_of(groups))) %>%
      mutate(
        residual = {
          model <- lm(reformulate(x, y), data = cur_data())
          preds <- predict(model, cur_data())
          (cur_data()[[y]] - preds) / cur_data()[[y]]
        },
        residual_class = factor(
          classify_residual(residual, threshold),
          levels = c("Below Expected", "Around Expected", "Above Expected")
          ) 
      )%>%
      ungroup()
  }

  df
}


residual_dot_plot <- function(
  residuals_df,
  x,
  y = "residual",
  color = "Attrition",
  threshold = 0.2,
  xlab = NULL,
  ylab = "Residual (ratio to actual income)",
  title = "Relative Pay vs. Time at Company"
) {
  
  # Generates a presentation ready dot plot for showing attrition in relation
  # to income expectations.

  ggplot(
    residuals_df,
    aes(x = .data[[x]], y = .data[[y]], color = .data[[color]])
  ) +
    # --- shaded regions ---
    annotate(
      "rect",
      xmin = -Inf, xmax = Inf, ymin = threshold, ymax = Inf,
      fill = green, alpha = 0.08
    ) +
    annotate(
      "rect",
      xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = -threshold,
      fill = magenta, alpha = 0.08
    ) +

    # --- points and reference lines ---
    geom_jitter(size = 1) +
    geom_hline(yintercept = threshold, linetype = "dashed", color = green) +
    geom_hline(yintercept = -threshold, linetype = "dashed", color = magenta) +

    # --- colors and theme ---
    scale_color_manual(
      values = c(
        "No" = scales::alpha(base1, 0.2),
        "Yes" = red
      )
    ) +
    labs(
      x = ifelse(is.null(xlab), x, xlab),
      y = ylab,
      title = title
    ) +
    ggthemes::theme_solarized(base_size = 18)
}

plot_attrition_heatmap <- function(
    residual_df, 
    x_var,
    y_var,
    title = "Heatmap of Attrition Rate",
    xlab = NULL,
    ylab = NULL
) {
  
  # Probes the distribution of attrition across two selected categorical
  # features.
  
  # Convert variable inputs to symbols for tidy evaluation
  x_sym <- ensym(x_var)
  y_sym <- ensym(y_var)
  
  # Default labels
  if (is.null(xlab)) xlab <- as_string(x_sym)
  if (is.null(ylab)) ylab <- as_string(y_sym)
  
  # Compute attrition rate by x × y
  attrition_summary <- residual_df %>%
    group_by(!!x_sym, !!y_sym) %>%
    summarise(
      n = n(),
      attrition_rate = mean(Attrition == "Yes", na.rm = TRUE) * 100,
      .groups = "drop"
    )
  
  # Plot heatmap
  p <- ggplot(attrition_summary, aes(x = !!x_sym, y = !!y_sym, fill = attrition_rate)) +
    geom_tile(color = base3, width = 0.9, height = 0.9) +
    geom_text(aes(label = sprintf("%.1f%%", attrition_rate)), color = "black", size = 6) +
    scale_fill_gradient(low = base3, high = cyan, name = "Attrition Rate (%)") +
    labs(
      title = title,
      x = xlab,
      y = ylab
    ) +
    theme_solarized(base_size = 18) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
  
  return(p)
}


quality_of_life_plot <- function(
  df,
  x = "Attrition",
  fill = "Attrition",
  facet_row = "WorkLifeBalance",
  facet_col = "OverTime",
  colors = c("No" = cyan, "Yes" = red),
  title = "Attrition vs. Quality of Life",
  subtitle = "For all employees",
  xlab = NULL,
  ylab = NULL
) {
  # Default labels if not provided
  if (is.null(xlab)) xlab <- facet_col
  if (is.null(ylab)) ylab <- facet_row
  
  ggplot(df, aes_string(x = x, fill = fill)) +
    geom_bar() +
    geom_text(
      stat = "count",
      aes(label = ..count..),
      vjust = -0.5
    ) +
    scale_fill_manual(values = colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
    facet_grid(
      as.formula(paste(facet_row, "~", facet_col)),
      switch = "both"
    ) +
    theme_solarized(base_size = 16) +
    labs(
      title = title,
      subtitle = subtitle,
      fill = fill,
      x = xlab,
      y = ylab
    ) +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      strip.placement = "outside"
    )
}
```

---------------------------------
## WHO IS ATTRITING?
---------------------------------
```{r}

# We start with the big pictures questions. What percentage of employees 
# are attriting? Which deparetments? Which Job Levels? How much does this all
# cost?

data %>% 
  group_by(Attrition) %>% 
  summarize(Count = n()) %>%
  mutate(Proportion = Count / sum(Count)) %>% 
  ggplot(aes(x = "", y = Proportion, fill = Attrition)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(
    aes(label = paste0(round(Proportion * 100, 1), "%")), 
    position = position_stack(vjust = 0.5),
    color = base3, size = 6, fontface = "bold"
  ) +
  scale_fill_manual(values = c("No" = cyan, "Yes" = red)) +
  labs(
    title = "Overall Attrition Rate",
    subtitle = paste0("Total Employees: ", nrow(data)),
    fill = "Attrition"
  ) +
  theme_solarized(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
ggsave("Project/01_pie_chart_of_total_attrition.png", create.dir = TRUE, width = 6, height =6)



DepartmentAttrition <- summarise_attrition(data, Department)
gtsave(
  render_attrition_table(DepartmentAttrition, Department),
  "Project/02_summary_of_department_attrition.png"
)


DepartmentWaste <- summarise_waste(data, Department)
gtsave(
  render_waste_table(DepartmentWaste, Department),
  "Project/03_summary_of_department_attrition.png"
)

data %>%
  filter(Attrition == "Yes") %>% 
  group_by(JobLevel, Department) %>% 
  summarize(Attrition_Count = n()) %>%
  ungroup() %>%
  group_by(Department) %>%
  mutate(Total_Attrition_in_Department = sum(Attrition_Count)) %>%
  mutate(Percent_Attrition = (Attrition_Count / Total_Attrition_in_Department) * 100) %>%
  select(Department, JobLevel, Attrition_Count, Percent_Attrition) %>% 
  ggplot(aes(x = factor(JobLevel), y = Department, fill = Percent_Attrition)) +
  geom_tile(color = base3) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Attrition)), color = "black", size = 8, fontface = "bold") +
  scale_fill_gradient(low = base3, high = cyan, name = "Attrition (%)") +
  labs(
    title = "Attrition Heatmap",
    x = "Job Level",
    y = "Department"
  ) +
  theme_solarized(base_size = 20) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(face = "bold")
  )
ggsave("Project/04_heatmap_of_attrition_by_department_and_level.png", width = 10, height =8)

# Process and render tables that summarises waste and attrition by Job Level

JobLevelAttrition <- summarise_attrition(data, JobLevel, arrange_col = JobLevel, descending = FALSE )
gtsave(
  render_attrition_table(JobLevelAttrition, JobLevel, group_label = "Job Level"),
  "Project/05_summary_of_job_level_attrition.png"
)

JobLevelWaste <- summarise_waste(data, JobLevel, arrange_col = JobLevel, descending = FALSE)
gtsave(
  render_waste_table(JobLevelWaste, JobLevel, group_label = "Job Level"),
  "Project/06_summary_of_job_level_waste.png"
)

data %>% 
  filter(Attrition == "Yes") %>% 
  ggplot(aes(x = YearsAtCompany, y = Department, fill = Department)) +
  geom_density_ridges() +
  theme_solarized(base_size = 20) +
  labs(
    title = "Density Plot of Attrition by Years at Frito-Lay",
    x = "Years at Company"
  ) +
  theme (
    axis.title.y = element_blank(),
    legend.position = "none"
    
  )
ggsave("Project/07_density_plot_of_attrition_by_years.png", width = 10, height =8)


```
---------------------------------
## WHY DO EMPLOYEES LEAVE
---------------------------------

```{r}

# This trio plots probes attrition in relation to two important factors:
# Work Life Balance and Overtime.

quality_of_life_plot(
  data, 
  xlab = "Overtime",
  ylab = "Work Life Balance"
  )
ggsave("Project/07_QoL_all.png", width =9, height = 6)

# The next two isolate lower and upper level employees, showing that
# quality of life is more important for lower level employees.

quality_of_life_plot(
  data %>% filter(JobLevel %in% c("Entry", "Junior")), 
  xlab = "Overtime",
  ylab = "Work Life Balance",
  subtitle = "For Entry and Junior Employees"
  )
ggsave("Project/08_QoL_jr.png", width =9, height = 6)

quality_of_life_plot(
  data %>% filter(!JobLevel %in% c("Entry", "Junior")), 
  xlab = "Overtime",
  ylab = "Work Life Balance",
  subtitle = "For Mid, Senior, and Excecutive Employees"
  )
ggsave("Project/09_QoL_sr.png", width =9, height = 6)

# This next investigation focuses on income. It calculates a residual that
# computes expected income as a function of `YearsSinceLastPromotion`, JobLevel,
# and Department.
#
# After computing residuals, its more clear that lower paid employees are more
# likely to leave.

residuals <- calc_residuals(
  data,
  "YearsSinceLastPromotion",
  "MonthlyIncome",
  groups = c("JobLevel", "Department"),
  threshold = 0.15
  )

p1 <- residual_dot_plot(
  residuals,
  "YearsSinceLastPromotion",
  xlab = "Years Since Last Promotion",
  title = "Income Residuals vs. Years at Company",
  threshold = 0.15
)

p2 <- plot_attrition_heatmap(
  residuals,
  "residual_class",
  "JobLevel",
  title = "Attrition Rate across Income Residuals",
  xlab = "Actual vs. Expected Income",
  ylab = "Job Level"
)

p1 / p2
ggsave("Project/10_income_residuals.png", width =6, height = 10)

# Probes the impact of another important reward, StockOptionLevel.
#
# Interesting, the impact was not linear. Groups with None and High Stock Options
# were more likely to attrite, while the other levels leave a ta.lower rate.

p3 <- plot_attrition_heatmap(
  residuals,
  "StockOptionLevel",
  "JobLevel",
  title = "Attrition Rate by Stock Option Level",
  xlab = "Stock Option Level",
  ylab = "Job Level"
)

p4 <- plot_attrition_heatmap(
  residuals %>% filter(residual_class == "Above Expected"),
  "StockOptionLevel",
  "JobLevel",
  title = "Only Highly Paid Employees",
  xlab = "Stock Option Level",
  ylab = "Job Level"
)

p3 / p4

ggsave("Project/11_stock_option_heatmap.png", width =6, height = 10)
```

---------------------------------
## MODEL EVALUATIONS
---------------------------------

```{r}
# Final Evalation that compares the models. Most graphics are 
# generated in the pipeline. A final evaluation compares metrics of
# the trial results.

nb_features <- all_of(c(KYE, Reward, CareerProgression, QualityOfLife))
nb_results <- model_eval_circuit(
  data,
  features = nb_features,
  n_trials = 1000
)

names(nb_results)
nb_results$impact_table
nb_results$cm_plot

ggsave("Project/13_nb_cm_plot.png", width = 8, height = 8)
nb_results$sensitivity_matrix
ggsave("Project/14_nb_sensitivity_matrix.png", width = 8, height = 8)
gtsave(
  nb_results$impact_table,
  "Project/15_nb_cost_impact.png")


knn_features <-c(
    "MonthlyIncome",
    "JobRole",
    "OverTime",
    "WorkLifeBalance",
    "StockOptionLevel",
    "YearsWithCurrManager",
    "PercentSalaryHike",
    "RelationshipSatisfaction"
    )

knn_results <- model_eval_circuit(
  data,
  features = knn_features,
  engineered_features = c("DepartmentGroup"),
  model_class = "knn",
  n_trials = 1000,
  balance = TRUE,
  group_col = "DepartmentGroup"
)

knn_results$cm_plot
knn_results$sensitivity_matrix
knn_results$impact_table

knn_results$cm_plot
ggsave("Project/16_knn_cm_plot.png", width = 8, height = 8)
knn_results$sensitivity_matrix
ggsave("Project/17_nb_sensitivity_matrix.png", width = 8, height = 8)
gtsave(
  knn_results$impact_table,
  "Project/18_knn_cost_impact.png")


nb_metrics <- nb_results$metrics
knn_metrics <- knn_results$metrics

# --- Combine model metrics ---
combined_metrics <- bind_rows(
  nb_metrics %>% mutate(Model = "Naive Bayes"),
  knn_metrics %>% mutate(Model = "k-NN")
)

# --- Compute density counts for each Model ---
heat_data <- combined_metrics %>%
  mutate(
    Sensitivity_bin = cut(Sensitivity, breaks = seq(0, 1, by = 0.05), include.lowest = TRUE),
    Specificity_bin = cut(Specificity, breaks = seq(0, 1, by = 0.05), include.lowest = TRUE)
  ) %>%
  group_by(Model, Sensitivity_bin, Specificity_bin) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(
    Sensitivity_mid = as.numeric(sub("\\((.+),.*", "\\1", Sensitivity_bin)) + 0.025,
    Specificity_mid = as.numeric(sub("\\((.+),.*", "\\1", Specificity_bin)) + 0.025
  )

# --- Plot heatmap ---
ggplot(heat_data, aes(x = Specificity_mid, y = Sensitivity_mid, fill = Count)) +
  geom_tile(color = "white", size = 0.2) +
  facet_wrap(~ Model) +
  scale_fill_viridis_c(option = "C", trans = "sqrt") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Model Performance Across 1000 Trials",
    x = "Specificity",
    y = "Sensitivity",
    fill = "N Trials"
  ) +
  theme_solarized(base_size = 18) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )
ggsave("Project/19_model_metrics_heatmap.png", width = 8, height = 4)
```
---------------------------------
## TEST SET PREDICTIONS
## (Brute Force Naive Bayes)
---------------------------------
```{r}
features = all_of(c(KYE, Reward, CareerProgression, QualityOfLife))

train <- data
test = read_csv(file.choose())

#Set Factors
test$BusinessTravel <- factor(
  test$BusinessTravel,
  levels = c("Non-Travel", "Travel_Rarely", "Travel_Frequently"),
  ordered = TRUE
)
test$Department <- as.factor(test$Department)
test$Education <- factor(
  test$Education,
  labels = c("Below College", "College", "Bachelor", "Master", "Doctor"),
  ordered = TRUE
)
test$EducationField <- as.factor(test$EducationField)
test$Gender <- as.factor(test$Gender)
test$JobInvolvement <- factor(
  test$JobInvolvement,
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
test$JobLevel <- factor(
  test$JobLevel,
  labels = c("Entry", "Junior", "Mid", "Senior", "Executive"),
  ordered = TRUE
)
test$JobRole <- as.factor(test$JobRole)
test$MaritalStatus <- as.factor(test$MaritalStatus)
test$Over18 <- as.factor(test$Over18)
test$OverTime <- as.factor(test$OverTime)
test$PerformanceRating <- factor(
  test$PerformanceRating,
  levels = c(1, 2, 3, 4),
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
test$RelationshipSatisfaction <- factor(
  test$RelationshipSatisfaction,
  labels = c("Low", "Medium", "High", "Very High"),
  ordered = TRUE
)
test$StockOptionLevel <- factor(
  test$StockOptionLevel,
  labels = c("None", "Low", "Medium", "High"),
  ordered = TRUE
)
test$WorkLifeBalance <- factor(
  test$StockOptionLevel,
  labels = c("Bad", "Good", "Better", "Best"),
  ordered = TRUE
)

splits <- list(
  train.X = train[ , features, drop = FALSE],
  train.Y = as.factor(train$Attrition),
  test.X  = test[ , features, drop = FALSE]
)
  
predictions <- get_naivebayes_predictions(splits)
output <- test %>% 
  select(ID) %>% 
  mutate(Attrition = predictions)
```







